### 2023-04-27 Manuscript/Draft1.Rmd
Manuscript update

The NEFSC has conducted a bottom trawl survey since 1963 to support assessment and management of the fish and invertebrate populations in the region [@azarovitz1981brief; @politis2014northeast]. The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf. The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits. The survey takes place twice each year- once in the spring and again in the fall. Since most spatial analyses and  projections of future distributions typically assume a constant survey catchability and/or availability over time, NOAA's survey design includes sampling Georges Bank during approximately the same 3-4 week time period in each season.

@@

Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades. The changes have resulted in a shift in spatial distributions of many species (cite Nye). Existing research has focused on temperature as the driver of such changes (cite klein, smith and kritzer) and evidence suggests that failing to account for the impact of climate-induced change can lead to management challenges (kerr, barajas, wiedenmann). Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments. More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey will change, altering the relationship between the index and the true population. We are therefore interested in analyzing the impact of climate change on the accuracy of future stock assessment models as measured by NOAA’s ongoing bottom-trawl survey along the East coast.

@@

To test the ability of the bottom trawl survey to track population trends under shifting environmental conditions, we construct spatial models for fish where movement depend on temperature preferences. We consider the impact of climate change by comparing simulations that use a repeating temperature pattern to those where temperature increases on average over time. In both cases we analyze the ability of stratified random sampling to track population trends through a design-based approach (stratified mean) compared to a model-based approach that allows for the inclusion of covariates (Vector Autoregressive Spatiotemporal Model).

@@

We construct spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock on Georges Bank, where movement of each species combine static species-specific habitat preferences with temperature preferences. Model dynamics are driven by a time series of temperature gradients that were estimated from data to create simulated data sets for each population where the true biomass is known. Using temperature gradients that repeat each year creates data sets with predictable, repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time. We conduct stratified random sampling on our simulation output to mimic the bottom trawl survey and use the samples to compare the ability of contemporary indexing methods to track population trends.

@@

The time step for our models is one week. MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass in each cell in our study area [@dolder2020highly]. Individual terms in the formulation account for growth of mature adults, natural and fishing mortality, and the addition of new recruits. Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period. Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics. A full list of parameters used in our model can be seen below in Tables \@ref(paramsALL) and \@ref(tab:paramsSCENARIOS).

@@

Species-specific habitat preferences were derived from niche model for each species using the *lrren* tool from the R package *envi* [@envi]. The *lrren* tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors. We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow. We combined data from both the fall and spring surveys to obscure the influence of temperature so that the niche model would instead infer habitat preferences. Depth and mean sediment size were used as our covariate predictors. Estimated depth for the region was obtained from FVCOM [@chen2006unstructured]. The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method **cite arcmap** using point data collected by the United States Geologic Survey (USGS) [@mcmullen2005usgs]. Since the values in $Hab_{J,s}$ are required to be between 0 and 1, we transform the spatial estimates from *lrren* to fall between these bounds. See Figure \@ref(fig:hab-plot1) for a visual representation of this process being applied to Cod. Figure \@ref(fig:hab-plot) depicts habitat preferences $Hab_{J,s}$ for each species.

@@

```{r hab-plot, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Static habitat preferences for each species in our population models (Yellotwtail, Cod, Haddock)."}
knitr::include_graphics(c("Images/Habitat_3species.png"))
```

@@

Each species is assumed to have normally distributed temperature preferences ($N(\mu,\sigma)$). Values were chosen by combining information in the literature with temperatures recorded in the bottom trawl survey. We assume Yellowtail Flounder's preferences are $N(8.75,4.25)$, while Haddock and Cod have preferences $N(9,4)$. Weekly estimated temperature data for the region for 2012 was obtained from FVCOM [@chen2006unstructured]. We chose to repeat estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data. The 2012 data was chosen because it displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values, allowing for a smooth repeating yearly temperature pattern for the constant temperature scenario. The 2012 temperature data was also transformed to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation. We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on dynamics while remaining within reasonable computational limits in terms of the length of the simulation. Figure \@ref(fig:temp-scenarios) depicts mean trends for the temperature scenarios used in our models. **dont forget to include animated gif in final submission**

@@

In equation (1), $Hab^2_{J,s}$ is constant for the duration of the simulation, while $Tol_{c,s,wk}$ changes each week. Using a temperature gradient that repeats every 52 weeks produces the same spatial preferences in a given week each year, resulting in consistent spatial biomass patterns. Scenarios where the temperature increases over time creates spatial preferences that evolve as the water warms, producing spatial biomass patterns that shift in a given week over the duration of the simulation. Thus, stratified random samples in scenarios with a repeating temperature pattern will have constant survey catchability and availability over time, which may not be true for increasing temperature scenarios due to evolving spatial preferences.

@@

After each simulation is complete, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year. We sample each strata in the same weeks in which the Spring and Fall surveys take place, and the number of the samples taken reflect true target values for each strata. Randomly selected cells were samples each year. We then use the biomass collected from our samples in contemporary abundance indexing methods to estimate population trends. Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.

@@

Stock assessment scientists use samples from the bottom trawl survey to estimate the abundance for each fish stock. There are a number of approaches to obtain abundance estimates including traditional design-based estimates to model-based estimates that range in complexity. Design-based estimators rely on the design of the sampling scheme with the underlying assumption that the data being collected is representative of the population of interest. These methods do not account for spatial variation in samples or allow the inclusion of environmental influences. Model-based abundance estimates use statistical models to measure the relationship between response variables (such as presence or abundance) and predictor variables (such as environmental factors). Model-based estimators, such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM), help account for complex relationships between variables and can help overcome problems with sampling design.

@@

We follow the advice given in [@thorson2019guidance] to build VAST models to estimate biomass on Georges Bank using stratified random samples from our model output. In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation. We also carried out the same process running models without covariate information as well as including covariates in our model. We considered covariates in the form of dynamic temperature values and/or static habitat values ($Hab_{J.s}$) from our population model. When using covariates we ultimately decided to provide the most information to the model by including both temperature and habitat covariates for both linear predictors. Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings. Through this process, and in consultation with the VAST package creator, we ultimately compared the performance of two sets of settings in our VAST models, which can be seen in Table\@ref(tab:VASTsettings).

Our goal is to determine indexing approaches and settings that are robust to future environmental conditions and resulting spatial biomass patterns. An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists. These assumptions can be questioned given enough noise in the sampling process **cite?** and/or climate change causing a population to move into previously uninhabited strata. To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values. **BL: Help with correct notation for adding noise**. We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species. We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from. Tabl XXX [[needs to be added]] lists all strata inhabited by each species, those that are removed from certain calculations, and the explanation of why these strata were removed.

<!-- Note the Table XXX needs to be added comment above. -->
When combining population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, covariates) there are a large number of scenario combinations to consider. The columns in Table \@ref(tab:scenarios) show the choices that define each scenario.

@@

The goal of our project was to analyze the ability of contemporary population indexing methods to track population trends under a variety of conditions, as depicted in Table \@ref(tab:scenarios). See Figure \@ref(fig:strata-plot) for a spatial reference of the Georges Bank strata and Figure \@ref(fig:hab-plot) for the static habitat preferences for each species ($Hab^2_{J,s}$) used in our population models.

Figure \@ref(fig:PopPct) depicts the spatial shifting that occurs in each stratum within our population models, specifically during the bottom trawl survey weeks in the spring (13 and 14) and fall (37 and 28). The left column depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern. In these scenarios we notice a small amount of shifting between successive years as the population aggregates on especially suitable habitat in the domain. More exaggerated shifting takes place in a larger number of strata when the temperature is increasing over time, as seen right column of Figure \@ref(fig:PopPct). This figure demonstrates how we were able to mimic shifting distributions due to climate change by running simulations with a temperature gradient that increased on average over time.

Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the results of our comparison between the absolute error between abundance estimates and model output. While the model-based results tended to provide a lower relative error to the stratified mean with a mean relative error of 0.34 compared to 0.38, the model-based estimates were very sensitive to the settings used. As a result, although $77/80\approx 96\%$ of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the variance for the VAST relative errors was much larger than the stratified mean (0.10 compared to 0.02). VAST models that include covariate information provided the lowest overall errors and standard deviations. When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal. Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain.

@@

The top two panels in Figure \@ref(fig:PopPct) depict the results for Yellowtail Flounder with a repeating temperature gradient on the left (constant temperature) and a temperature gradient that increases over time on the right (increasing temperature). In both temperature scenarios we see the percent of Yellowtail Flounder in strata 13 decrease over the course of the time series in both seasons. The spring population in stratum 19 also decreases in both temperature scenarios as well. The percent of the population increases in stratum 16 in the spring over the duration of both time series, which implies the flow out of strata 13 and 19 in the spring are going into stratum 16. These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail Flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12. These spring dynamics are therefore related to the static habitat values in our model rather than the dynamics temperature preferences, wihch is why we seen the same dynamics with constant and increasing temperature. While we observe similar dynamics in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \@ref(fig:PopPct)). These dynamics imply that an increase in temperature results in the more shallow strata 16 becoming less desirable than the deeper and more narrow exterior strata 17 and 18. One noticeable seasonal difference in the constant temperature scenario for Yellowtail is how ~10\% of the population exists in the narrow exterior strata 17 in the fall, while seemingly none of the population exists in any of the exterior strata (14, 15, 17, 18) in the spring.

In considering the Yellowtail Flounder results in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata), we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors. When all strata are sampled, adding covariates greatly improved estimates in all scenarios. We still see an improvement in VAST estimates when certain strata are excluded from sampling, but the improvement was much less dramatic than with all strata included. However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata). Including covariate information to each VAST model produced estimates with significantly lower error than their stratified mean counterparts.

Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata) contain the absolute error between our abundance estimates for Yellowtail Flounder comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B). Of all scenarios without covariates, $33/48\approx68$\% had a VAST fit with a lower relative error compared to the stratified mean estimate. All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the fall season and had a common theme of producing abundance estimates that are above the true model value. These 15 fall estimates span all other scenario variations. The implication of this is that our model-based approach without covariates struggles with the primary seasonal difference for Yellowtail Flounder, which is that a larger percentage of the population exists in the narrow exterior strata (18 and/or 17). This theory is further supported by the fact that the absolute error in the increasing temperature scenarios increased dramatically in the fall season, when the combined percentage of the population in the outer strata 17 and 18 increased to over 40\% by the end of the simulation.

@@

<font size="5">*Model / Estimate Ratio Results*</font>

A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, too low. This can be further examined by viewing estimate/model ratio plots, where we divide the yearly abundance estimates by the true model value and plot them on the log scale. In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can range as low as 0 and exhibit large yearly changes. A representative example of this trend is shown in Figure \@ref(fig:RatioTrend)

A more careful analysis of individual yearly model/estimate ratios when all strata are included revealed that about 73\% of all VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21. On the other hand, 33\% of stratified mean estimates were above 1 with an average of 0.874 and a standard deviation of 0.12. There were seasonal differences in estimate ratios with spring VAST ratios producing (1.08,0.12), fall VAST ratios being (1.50,0.38), stratified mean spring ratios resulting in (0.91,0.10), and fall values of (0.84,0.16). The breakdown for individual species followed a similar pattern.

When the entire domain is sampled, adding covariates brings the estimate ratio closer to 1. However, when failing to sample the entire domain, adding covariates can decrease the accuracy of VAST estimates. When including all strata, the stratified mean produced (0.87,0.12), while adding covariates improved the VAST estimate ratio from (1.45,0.17) to (1.13,0.09). With a reduced number of strata, the stratified mean resulted in our worst summary values of (0.59,0.10), and adding covariates to our VAST models decreased the average ratio results from (1.04,0.33 ) to (0.78,0.19).

@@

As see in Table **cite VAST setting table**, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses. While using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates with $77/80\approx 96\%$ of VAST models with covariates providing lower absolute error than the corresponding stratified mean estimate, there is a great deal more to explore with respect to the impact of covariates. For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates. Adding noise to our covariate input would test how robust the model-based estimates are to uncertainty in the covariate information. One could also test the impact of assuming the wrong covarite response function (linear vs normally distributed etc). 
**For discussion: clearly our perfect covariate information made a big difference for yellowtail flounder (and others), but the question is could we achieve similar results without "perfect" information? Which covariate is best? Good question for future research**

>>>>

The NEFSC has conducted a bottom trawl survey since 1963 to support assessment and management of the fish and invertebrate populations in the region [@azarovitz1981brief; @politis2014northeast]. The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf. The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits. The survey takes place twice each year- once in the spring and again in the fall. Most spatial analyses and  projections of future distributions typically assume a constant survey catchability and/or availability over time. For this reason, NOAA's survey design includes sampling each strata in approximately the same 3-4 week time period in each season.

@@

Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades. The changes have resulted in a shift in spatial distributions of many species [@nye2009changing; @kleisner2017marine; @henderson2017effects]. Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments. More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey will change, altering the relationship between the index and the true population. A species shifting its range beyond the survey area is an additional compounding factor to consider. Existing research has focused on temperature as the driver of such changes [@klein2017effects] and evidence suggests that failing to account for the impact of climate-induced change can lead to management challenges [@kerr2022coherence]. In these scenarios, management strategy evaluations have shown that unintended overfishing can occur resulting from misconceptions of stock status, which can ultimately have detrimental ecologic and economic impacts [@mazur2023consequences]. We are therefore interested in analyzing the impact of climate change on the accuracy of abundance estimates derived from NOAA’s ongoing bottom-trawl survey along the East coast.

@@

To test the ability of the bottom trawl survey to track population trends under shifting environmental conditions, we construct spatial models for fish where movement depend on temperature preferences. We consider the impact of climate change by comparing simulations that use a repeating water temperature pattern to those where temperature increases on average over time. In both cases we analyze the ability of stratified random sampling to track population trends through a design-based approach (stratified mean) compared to a model-based approach that allows for the inclusion of covariates (Vector Autoregressive Spatiotemporal Model).

@@

We construct spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock on Georges Bank, where movement of each species combine static species-specific habitat preferences with biologically-based temperature preferences. Model dynamics are driven by a time series of temperature gradients to create simulated data sets for each population where the true biomass is known. Using temperature gradients that repeat each year creates data sets with repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time. We conduct stratified random sampling on our simulation output to mimic the bottom trawl survey and use the samples to compare the ability of contemporary indexing methods to track population trends.

@@

The time step for our models is one week. MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass in each cell in our study area [@dolder2020highly]. Individual terms in the formulation account for growth of mature adults, mortality (natural and fishing), and the addition of new recruits. Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period. Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics. A full list of parameters used in our model can be seen below in Tables \@ref(paramsALL) and \@ref(tab:paramsSCENARIOS).

@@

Species-specific habitat preferences were derived from niche model for each species using the *lrren* tool from the R package *envi* [@envi]. The *lrren* tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors. We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow. We combined data from both the fall and spring surveys to obscure the influence of temperature to allow the niche model to instead infer habitat preferences. Depth and mean sediment size were used as our covariate predictors. Estimated depth for the region was obtained from FVCOM [@chen2006unstructured]. The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method using point data collected by the United States Geologic Survey (USGS) [@mcmullen2005usgs]. Since the values in $Hab_{J,s}$ are required to be between 0 and 1, we rescaled the spatial estimates from *lrren* to fall between these bounds. See Figure \@ref(fig:hab-plot1) for a visual representation of this process being applied to Cod. Figure \@ref(fig:hab-plot) depicts habitat preferences $Hab_{J,s}$ for each species.

@@

```{r hab-plot, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Static habitat preferences for each species in our population models. From left to right: Yellotwtail Flounder, Cod, and Haddock."}
knitr::include_graphics(c("Images/Habitat_3species.png"))
```

@@

Each species is assumed to have normally distributed temperature preferences ($N(\mu,\sigma)$). Values were chosen by combining information in the literature with temperatures recorded in the bottom trawl survey. We assume Yellowtail Flounder's preferences are $N(8.75,4.25)$, while Haddock and Cod have preferences $N(9,4)$. Weekly estimated temperature data for the region for 2012 was obtained from FVCOM [@chen2006unstructured]. We chose to repeat temperature estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data. The 2012 data was chosen because it displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values, allowing for a smooth repeating yearly temperature pattern for the constant temperature scenario. The 2012 temperature data was also transformed to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation. We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on dynamics while remaining within reasonable computational limits in terms of the length of the simulation. Figure \@ref(fig:temp-scenarios) depicts mean trends for the temperature scenarios used in our models. **dont forget to include animated gif in final submission**

@@

In equation (1), $Hab^2_{J,s}$ remains constant for each species for the duration of the simulation, while $Tol_{c,s,wk}$ changes each week with temperature fluctuations. Using a temperature gradient that repeats every 52 weeks produces the same spatial preferences in a given week each year, resulting in consistent spatial biomass patterns. Scenarios where the temperature increases over time creates spatial preferences that evolve as the water warms, producing spatial biomass patterns that shift in a given week over the duration of the simulation. Thus, stratified random samples in scenarios with a repeating temperature pattern will have constant survey catchability and availability over time, which may not be true for increasing temperature scenarios due to evolving spatial preferences.

@@

After each simulation is complete, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year. We sample each strata in the same weeks in which the Spring and Fall surveys take place (weeks 13 & 14 in the Spring and 37 & 38 in the Fall). The number of the samples taken reflect true target values for each strata and sampling cells were randomly selected. We then use the biomass collected from our samples in contemporary abundance indexing methods to estimate population trends. Knowing the true population values in our simulations allows us to compute and compare the absolute error calculated from each estimation method.

@@

Stock assessment scientists choose from a number of approaches to obtain abundance estimates that are derived from the survey data. Methods range from traditional design-based estimates to model-based estimates that vary in complexity. Design-based estimators rely on the design of the sampling scheme with the underlying assumption that the data being collected is representative of the population of interest. These methods do not account for spatial variation in samples and are not able to account for environmental influences on survey values. Model-based abundance estimates use statistical models to measure the relationship between response variables (such as presence or abundance) and predictor variables (such as environmental factors). Model-based estimators, such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM), help account for complex relationships between variables and can help overcome problems with sampling design.

@@

We follow the advice given in [@thorson2019guidance] to build VAST models to estimate biomass in our Georges Bank population models using stratified random samples from our model output. In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation. We carried out the same model-building process running models without covariate information as well as including covariates in our model. We considered covariates in the form of dynamic temperature values and/or static habitat values ($Hab_{J.s}$) from our population models. When using covariates we ultimately decided to consider a best-case scenario by including both temperature and habitat covariates for both linear predictors in order to provide the most information to the model. Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings. Through this process, and in consultation with the VAST package creator, we ultimately compared the performance of two sets of settings in our VAST models, which can be seen in Table\@ref(tab:VASTsettings).

Our goal is to determine indexing approaches and settings that are robust to changing environmental conditions and resulting spatial biomass patterns. An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists. These assumptions can be questioned given enough noise in the sampling process **cite?** and/or climate change causing a population to move into previously uninhabited strata. To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values. **BL: Help with correct notation for adding noise**. The impication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population. We therefore simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species. We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from. Tabl XXX [[needs to be added]] lists all strata inhabited by each species, those that are removed from certain calculations, and the explanation of why these strata were removed. The yellow regions in Figure \@ref(fig:strata-plot) depict the strata that were removed from certain stratified random surveys for each species.

<!-- Note the Table XXX needs to be added comment above. -->
Each scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, covariates), resulting in a large number of scenario combinations to consider. The columns in Table \@ref(tab:scenarios) show the choices that define each scenario.

@@

The goal of our project was to analyze the ability of contemporary population indexing methods to track population trends under the variety of conditions shown in Table \@ref(tab:scenarios). and Figure \@ref(fig:hab-plot) for the static habitat preferences for each species ($Hab^2_{J,s}$) used in our population models.

Figure \@ref(fig:PopPct) depicts the spatial shifting that occurs in each stratum within our population models, specifically during the bottom trawl survey weeks in the spring (13 and 14) and fall (37 and 28). See Figure \@ref(fig:strata-plot) for a spatial reference of the Georges Bank strata. The left column in Figure \@ref(fig:PopPct) depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern. In these scenarios we notice a small amount of shifting between successive years as the population aggregates on especially suitable habitat in the domain. More exaggerated shifting takes place in a larger number of strata when the temperature is increasing over time, as seen right column of Figure \@ref(fig:PopPct). This demonstrates how we were able to model shifting distributions due to climate change by running simulations with a temperature gradient that increased on average over time.

Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the absolute error between abundance estimates and model output for each of our abundance estimates. While the model-based results tended to provide a slightly lower mean absolute error compared to the stratified mean with a mean relative error of 0.34 compared to 0.38, the variance seen in VAST relative errors was much larger than the stratified mean (0.10 compared to 0.02). VAST models that include covariate information provided the lowest overall errors and standard deviations. As a result, while $77/80\approx 96\%$ of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the high variance in error values indicate how sensitive model-based estimates were to the settings used. When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal. Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain.

@@

The top two panels in Figure \@ref(fig:PopPct) depict the results for Yellowtail Flounder with a repeating temperature gradient on the left (constant temperature) and a temperature gradient that increases over time on the right (increasing temperature). In both temperature scenarios we see the percent of Yellowtail Flounder in strata 13 decrease over the course of the time series in both seasons. The spring population in stratum 19 also decreases in both temperature scenarios as well. The percent of the population increases in stratum 16 in the spring over the duration of both time series, which implies the flow out of strata 13 and 19 in the spring are going into stratum 16. These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail Flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12. These spring dynamics are therefore related to the static habitat values in our model rather than the dynamics temperature preferences, which is why we seen the same dynamics with constant and increasing temperature. While we observe similar dynamics in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \@ref(fig:PopPct)). These dynamics imply that an increase in temperature results in the more shallow strata 16 becoming less desirable than the deeper and more narrow exterior strata 17 and 18. One noticeable seasonal difference in the constant temperature scenario for Yellowtail is how ~10\% of the population exists in the narrow exterior strata 17 in the fall, while seemingly none of the population exists in any of the exterior strata (14, 15, 17, 18) in the spring.

Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata) contain the absolute error between our abundance estimates for Yellowtail Flounder comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B). In reviewing these Tables we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors. When all strata are sampled, adding covariates greatly improved estimates in all scenarios. We still see an improvement in VAST estimates when certain strata are excluded from sampling, but the improvement was much less dramatic than with all strata included. There are several instances in which VAST failed to provide improved abundance estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata). However, Including covariate information to each VAST model produced estimates with significantly lower error than their stratified mean counterparts.

Of all scenarios without covariates, $33/48\approx68$\% had a VAST fit with a lower relative error compared to the stratified mean estimate. All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the fall season and had a common theme of producing abundance estimates that are above the true model value. These 15 fall estimates span all other scenario variations. The implication of this is that our model-based approach without covariates struggles with the primary seasonal difference for Yellowtail Flounder, which is that a larger percentage of the population exists in the narrow exterior strata (18 and/or 17). This theory is further supported by the fact that the absolute error in the increasing temperature scenarios increased dramatically in the fall season, when the combined percentage of the population in the outer strata 17 and 18 increased to over 40\% by the end of the simulation.

@@

<font size="5">*Abundance Estimate Ratio Results*</font>

A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values. This can be further examined by viewing yearly estimate/model ratio plots, where we divide the yearly abundance estimates by the true model value and plot them on the log scale. In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can range as low as 0 and exhibit large yearly changes. A representative example of this trend is shown in Figure \@ref(fig:RatioTrend).

Analysis of individual yearly model/estimate ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21. On the other hand, 33\% of stratified mean estimates were above 1 with an average of 0.874 and a standard deviation of 0.12. There were seasonal differences in estimate ratios for VAST with spring VAST ratios producing a mean value of 1.08 with a standard deviation of 0.12 while fall VAST ratios were larger with a mean of 1.50 and standard deviation of 0.38. Stratified mean ratios were more consistent between seasons with spring ratios resulting in a mean value of 0.91 with a standard deviation of0.10, and fall values of providing a mean ratio of 0.84 with a standard deviation of 0.16. The breakdown for individual species followed a similar pattern.

When the entire domain is sampled, adding covariates brings the estimate ratio closer to 1. Specifically, when including all strata, the stratified mean produced an average yearly ration of 0.87 with a standard deviation of 0.12, while adding covariates improved the VAST estimate ratio from a mean of 1.45 and standard deviation of 0.17 to a mean of 1.13 and standard deviation of 0.09.  However, when failing to sample the entire domain, adding covariates can decrease the accuracy of VAST estimates. With a reduced number of strata, the stratified mean resulted in our worst ratio estimate with a mean value of 0.59 and standard deviation of 0.10, and adding covariates to our VAST models decreased the average ratio results from a mean of 1.04 and standard devnation of 0.33 to a mean of 0.78 and standard deviation of 0.19.

@@

As see in Table **cite VAST setting table**, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses. While using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates with $77/80\approx 96\%$ of VAST models with covariates providing lower absolute error than the corresponding stratified mean estimate, there is more to explore with respect to the impact of covariates. For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates. Adding noise to our covariate input would test how robust the model-based estimates are to uncertainty in the covariate information. One could also test the impact of assuming the wrong covarite response function (linear vs normally distributed etc). 
<!-- **For discussion: clearly our perfect covariate information made a big difference for yellowtail flounder (and others), but the question is could we achieve similar results without "perfect" information? Which covariate is best? Good question for future research** -->

As noted in our results, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates, when a reduced number of strata are sampled adding covariate information leads to a decrease in performance for fall estimates of Haddock and Cod. The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, leading to an incomplete estimation of the covariate response. Figure \@ref(fig:strata-plot) shows the strata excluded from sampling and Figure \@ref(fig:CovResponse) depicts the resulting covariate response for Yellowtail Flounder and Haddock for the same population scenario. Although strata were excluded for both species, the survey samples for Yellowtail Flounder still ranged across the species preferred values and thus fully formed the covariate response, while the Haddock covariate response did not contain enough samples from the lower temperature range due to the specific strata that were excluded from sampling (ie, northern strata). As a result, when projecting into those strata as seen in the lower portion of Figure \@ref(fig:CovResponse), Yellowtail estimates with covariate information approximate the correct trend despite these strata being excluded from sampling, while the Haddock estimates that include strata provide that wrong trend and estimate the wrong magnitude of biomass.
